---
title: "Introduction to bmco"
output: rmarkdown::html_vignette
bibliography: bmco.bib
nocite: '@*'
vignette: >
  %\VignetteIndexEntry{Introduction to bmco}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
% Source - https://stackoverflow.com/a/61528568
% Posted by J_F
% Retrieved 2026-02-18, License - CC BY-SA 4.0

```{r, echo=FALSE}
biblio <- RefManageR::ReadBib("bmco.bib", check = "error")
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The **bmco** package provides Bayesian methods for analyzing multiple binary outcomes simultaneously. It addresses a common problem in clinical trials, educational research, and other fields: how to compare groups when success is defined by multiple endpoints.

### The Problem

Suppose you're evaluating a new drug and want to know if it improves **both** symptom relief and quality of life. 

`bmco` provides a principled Bayesian framework that:

- Analyzes **multiple binary outcomes jointly**
- Accounts for **correlations between outcomes**
- Offers **flexible decision rules** (All, Any, or weighted combinations)
- Provides **posterior probabilities** 
- Estimates **subgroup effects** using data from the entire sample
- Handles **clustered data**

## Quick Start Example

```{r setup}
library(bmco)
set.seed(2024)

# Simulate a clinical trial with two binary outcomes
trial_data <- data.frame(
  treatment = rep(c("placebo", "drug"), each = 50),
  symptom_relief = rbinom(100, 1, prob = rep(c(0.4, 0.6), each = 50)),
  quality_of_life = rbinom(100, 1, prob = rep(c(0.5, 0.7), each = 50))
)

# Test if drug improves BOTH outcomes
result <- bmvb(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  y_vars = c("symptom_relief", "quality_of_life"),
  test = "right_sided",
  rule = "All",
  n_it = 5000
)

print(result)
```

The posterior probability (`pop`) of this example tells us: **"What is the probability that the drug is better than placebo on BOTH outcomes?"**

## Three Analysis Functions

The package provides three main functions for different scenarios:

### 1. `bmvb()`: Basic Comparison

**Use when**: Comparing two groups on multiple binary outcomes. Currently supported for two outcomes only.

```{r bmvb_example, eval=FALSE}
# Compare two teaching methods on two outcomes
bmvb(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  y_vars = c("symptom_relief", "quality_of_life"),
  rule = "All",
  n_it = 5000
)
```

For more details on statistical methodology or when using this function, please cite:

`r capture.output(print(biblio["Kavelaars2020"]))`

### 2. `bglm()`: Subgroup analysis

**Use when**: Comparing two subgroups on multiple binary outcomes (e.g., age range, particular baseline severity). Currently supported for two outcomes only.

```{r bglm_example, eval=FALSE}
# Add age covariate
trial_data$age <- rnorm(100, mean = 50, sd = 10)

result_bglm <- bglm(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  x_var = "age",
  y_vars = c("symptom_relief", "quality_of_life"),
  x_method = "Empirical",
  x_def = c(-Inf, Inf),
  rule = "All",
  n_burn = 1000,
  n_it = 3000
)
```

For more details on statistical methodology or when using this function, please cite:

`r capture.output(print(biblio["Kavelaars2024"]))`

Also see the "Subgroup analysis" vignette for details.

### 3. `bglmm()`: With Clustering

**Use when**: Comparing two (sub)groups on multiple binary outcomes, when data are clustered (patients within hospitals, students within schools).

```{r bglmm_example, eval=FALSE}
# Add cluster variable
trial_data$hospital <- rep(1:10, each = 10)

result_bglmm <- bglmm(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  id_var = "hospital",
  x_var = "age",
  y_vars = c("symptom_relief", "quality_of_life"),
  x_method = "Empirical",
  x_def = c(-Inf, Inf),
  rule = "All",
  n_burn = 1000,
  n_it = 3000,
  n_thin = 3
)
```

For more details on statistical methodology or when using this function, please cite:

`r capture.output(print(biblio["Kavelaars2023"]))`

Also see the "Multilevel Models" vignette for details.

## Decision Rules

Choose how to define "treatment success":

### All Rule (Conjunctive)

Treatment must improve **ALL** outcomes:

```{r rule_all}
result_all <- bmvb(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  y_vars = c("symptom_relief", "quality_of_life"),
  rule = "All",  # BOTH outcomes must improve
  n_it = 5000
)

cat("P(drug better on BOTH outcomes):", result_all$delta$pop, "\n")
```

### Any Rule (Disjunctive)

Treatment must improve **AT LEAST ONE** outcome:

```{r rule_any}
result_any <- bmvb(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  y_vars = c("symptom_relief", "quality_of_life"),
  rule = "Any",  # At least one outcome must improve
  n_it = 5000
)

cat("P(drug better on AT LEAST ONE outcome):", result_any$delta$pop, "\n")
```

### Compensatory Rule (Weighted)

Weight outcomes by importance:

```{r rule_comp}
result_comp <- bmvb(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  y_vars = c("symptom_relief", "quality_of_life"),
  rule = "Comp",
  w = c(0.7, 0.3),  # Symptom relief weighted 70%, QoL 30%
  n_it = 5000
)

cat("P(weighted improvement):", result_comp$delta$pop, "\n")
cat("Weighted effect:", result_comp$delta$w_delta, "\n")
```

## Test Directions

### Right-sided Test

Test if group B is better than group A:

```{r test_right}
bmvb(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",      
  grp_b = "drug", # Put the group you expect to be better here
  y_vars = c("symptom_relief", "quality_of_life"),
  test = "right_sided",  # P(drug > placebo)
  n_it = 5000
)
```

### Left-sided Test

Test if group A is better than group B:

```{r test_left}
bmvb(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  y_vars = c("symptom_relief", "quality_of_life"),
  test = "left_sided",  # P(placebo > drug)
  n_it = 5000
)
```

**Note**: Left-sided with (A, B) is equivalent to right-sided with (B, A).

## Understanding the Output

```{r output_structure}
result <- bmvb(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  y_vars = c("symptom_relief", "quality_of_life"),
  n_it = 5000
)

# Key components
names(result)

# Posterior estimates for each group
result$estimates

# Sample sizes
result$sample_sizes

# Treatment effect
result$delta

# Test information
result$info
```

### Key Output Elements

- `estimates$mean_a` / `mean_b`: Posterior mean success probabilities for each group
- `estimates$sd_a` / `sd_b`: Posterior standard deviations
- `delta$mean_delta`: Mean treatment effect (difference in probabilities)
- `delta$se_delta`: Standard error of treatment effect
- `delta$pop`: **Posterior probability** (the main result!)
- `info$test_label`: Human-readable description of the test

## Posterior Samples

For custom analyses, extract posterior samples:

```{r samples}
result_samples <- bmvb(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  y_vars = c("symptom_relief", "quality_of_life"),
  n_it = 5000,
  return_samples = TRUE
)

# Access posterior samples
str(result_samples$samples)

# Custom probability calculations
samples <- result_samples$samples$delta

# P(effect on symptom relief > 0.1)
cat("P(delta[symptom] > 0.1):", mean(samples[, 1] > 0.1), "\n")

# P(effect on QoL > 0.15)
cat("P(delta[QoL] > 0.15):", mean(samples[, 2] > 0.15), "\n")

# Credible intervals
cat("\n95% Credible Intervals:\n")
print(apply(samples, 2, quantile, probs = c(0.025, 0.975)))
```

## Practical Considerations

### Sample Size

For basic comparison: see @Kavelaars2020. 
For subgroup analysis: see @Kavelaars2024.

### MCMC Settings

Default settings usually work, but you can adjust:

```{r mcmc_settings, eval=FALSE}
bmvb(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  y_vars = c("symptom_relief", "quality_of_life"),
  n_it = 10000  # More iterations for smoother estimates
)
```

For `bglm()` and `bglmm()`:

```{r mcmc_settings_reg, eval=FALSE}
bglm(
  # ... arguments ...
  n_burn = 2000,  # Burn-in / warm-up iterations
  n_it = 5000,    # Sampling iterations
  n_thin = 1      # Keep every nth sample (reduces memory)
)
```

### Missing Data

Missing values are automatically handled via complete-case analysis:

```{r missing_data, eval=FALSE}
# Introduce missing data
trial_data_missing <- trial_data
trial_data_missing$symptom_relief[1:5] <- NA

result_missing <- bmvb(
  data = trial_data_missing,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  y_vars = c("symptom_relief", "quality_of_life"),
  n_it = 5000
)

# Sample sizes are reduced
result_missing$sample_sizes
```

## Comparison with Frequentist Approaches

### Traditional approach

```{r frequentist, eval=FALSE}
# Separate tests for each outcome
chisq_symptom <- chisq.test(table(trial_data$treatment, trial_data$symptom_relief))
chisq_qol <- chisq.test(table(trial_data$treatment, trial_data$quality_of_life))

cat("Frequentist p-values:\n")
cat("  Symptom relief:", chisq_symptom$p.value, "\n")
cat("  Quality of life:", chisq_qol$p.value, "\n")
cat("  Bonferroni-adjusted alpha: 0.025 (needed for Any-rule only; not needed for All-rule)\n\n")
```

### Bayesian approach

```{r bayesian, eval=FALSE}
result_bayes <- bmvb(
  data = trial_data,
  grp = "treatment",
  grp_a = "placebo",
  grp_b = "drug",
  y_vars = c("symptom_relief", "quality_of_life"),
  rule = "All",
  n_it = 5000
)

cat("Bayesian result:\n")
cat("  P(drug better on BOTH):", result_bayes$delta$pop, "\n")
```

### Advantages of Bayesian approach

1. **Direct probability statements**: "95% probability drug is better" vs. "reject null at alpha=0.05"
2. **Flexible decision rules**: All/Any/Weighted combinations
4. **Coherent uncertainty**: Full sample of posterior distribution available

## Next Steps

Explore the other vignettes:

- **Subgroup analysis**: Using `bglm()` to adjust for baseline variables
- **Multilevel Models**: Using `bglmm()` for clustered data

## References

**Statistical methodology**:
More details about statistical methodology can be found here in the papers below.

When using the `bmvb()` function, please cite:
`r capture.output(print(biblio["Kavelaars2020"], .opts = list(style = "markdown", bib.style = "authoryear", sorting = "ydnt")))`

When using the `bglm()` function, please cite:
`r capture.output(print(biblio["Kavelaars2024"], .opts = list(style = "markdown", bib.style = "authoryear", sorting = "ydnt")))`

When using the `bglmm()` function, please cite:
`r capture.output(print(biblio["Kavelaars2023"], .opts = list(style = "markdown", bib.style = "authoryear", sorting = "ydnt")))`
